---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  echo = TRUE
)
knitr::opts_knit$set(
  global.device = TRUE # shares graphic device between chunks
)
```

# raem

<!-- badges: start -->

<!-- badges: end -->

`raem` is an R package for modeling steady-state single-layer groundwater flow under the Dupuit-Forchheimer assumption using analytic elements.

## Installation

You can install the development version of `raem` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("cneyens/raem")
```

## Example

Construct an analytic element model of an aquifer with uniform background flow, two extraction wells and a reference point.

Specify the aquifer parameters and create elements:

```{r create-elements}
library(raem)

k = 10
top = 10
base = 0
n = 0.2

hr = 15 # head at reference point, above aquifer top
TR = k * (top - base) # constant transmissivity of background flow

uf = uniformflow(TR, gradient = 0.001, angle = -45)
rf = constant(xc = -1000, yc = 0, hc = hr)
w1 = well(xw = 200, yw = 0, Q = 250)
w2 = well(xw = -200, yw = -150, Q = 400)
```

Create the model. This automatically solves the system of equations.

```{r create-model}
m = aem(k = k, top = top, base = base, n = n, 
        uf, rf, w1, w2)
```

Find the head and discharge at two locations `x = -200, y = 200` and `x = 100, y = 200`. Note that there are no vertical flow components in this model:

```{r find-head}
heads(m, x = c(-200, 100), y = 200)

discharge(m, c(-200, 100), 200, z = top) # [L^2 / T]
```

Plot head contours and element locations. First, specify the contouring grid:

```{r set-grid}
xg = seq(-500, 500, length = 100)
yg = seq(-250, 250, length = 100)
```

Now plot:

```{r plot-head, dpi = 300}
contours(m, xg, yg, 'heads', col = 'dodgerblue', nlevels = 20)
plot(m, add = TRUE)
```

Compute particle traces starting along `y = 200` at 20 intervals per year for 5 years and add to plot:

```{r plot-traces, dpi = 300}
paths = tracelines(m, x0 = seq(-450, 450, 50), y0 = 200, z0 = top, times = seq(0, 5*365, 365/20))
plot(paths, add = TRUE, col = 'orange')
```

```{r turn-off-device, include = FALSE, eval = FALSE}
# turn off global graphics device otherwise plot is also shown in following chunks
# this still creates a plot in man/figures, so setting eval = FALSE
knitr::opts_knit$set(global.device = FALSE)
```
