---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  echo = TRUE
)
knitr::opts_knit$set(
  global.device = TRUE # shares graphic device between chunks
)
```

# raem

<!-- badges: start -->
<!-- badges: end -->

`raem` is an R package for modeling steady-state single-layer groundwater flow using analytic elements.

## Installation

You can install the development version of raem from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("cneyens/raem")
```

## Example

Construct an analytic element model with uniform background flow, two extraction wells and a reference point.

Specify the aquifer parameters and create elements:
```{r create-elements}
library(raem)

k = 10
top = 10
base = 0
n = 0.2
TR = k * (top - base) # transmissivity

uf = uniformflow(TR, gradient = 0.001, angle = -45)
rf = constant(xc = -1000, yc = 0, hc = 10)
w1 = well(xw = 200, yw = 0, Q = 250)
w2 = well(xw = -200, yw = -150, Q = 400)
```

Create the model. This automatically solves the system of equations.
```{r create-model}
m = aem(k = k, top = top, base = base, n = n, 
        uf, rf, w1, w2)
```

Find the head and discharge at two locations `x = 50, y = 200` and `x = 100, y = 200`:
```{r find-head}
heads(m, x = c(50, 100), y = 200)

discharge(m, c(50, 100), 200, z = top) # [L^2 / T]
```

Plot head contours and element locations. First, specify the contouring grid:
```{r set-grid}
xg = seq(-500, 500, length = 100)
yg = seq(-250, 250, length = 100)
```

Now plot:
```{r plot-head, dpi = 300}
contour(m, xg, yg, z = 'heads', col = 'dodgerblue3', nlevels = 20)
plot(m, add = TRUE)
```

Compute particle traces starting along `y = 200` at 20 intervals per year for 5 years and add to plot:
```{r plot-traces, dpi = 300}
paths = tracelines(m, x0 = seq(-450, 450, 50), y0 = 200, z0 = top, times = seq(0, 5*365, 365/20))
plot(paths, add = TRUE, col = 'orange3')
```

```{r turn-off-device, include = FALSE, eval = FALSE}
# turn off global graphics device otherwise plot is also shown in following chunks
# this still creates a plot in man/figures, so setting eval = FALSE
knitr::opts_knit$set(global.device = FALSE)
```

