---
title: "Exporting spatial data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exporting spatial data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, 
  fig.height = 5, 
  fig.align = "default"
)
```

`raem` does not contain functions to export output to spatial data formats such as vector data or rasters. Other R packages do have this functionality however. This vignette describes a way to create spatial data from `raem` output using other, well-established, packages.

# Create an example AEM model

As an example model, two extraction wells in a phreatic aquifer with uniform background flow and areal recharge are modeled. The output grid is set up and head contours are plotted.

```{r create-model}
library(raem)

# aquifer parameters
k = 10    # hydraulic conductivity, m/d
top = 30  # aquifer top elevation, m
base = 0  # aquifer bottom elevation, m
n = 0.2   # aquifer effective porosity, -

# create elements
w1 = well(x = -500, y = 100, Q = 1000)
w2 = well(x = 300, -200, Q = 1200)
as = areasink(x = 0, y = 0, R = 1500, N = 0.3 / 365)
uf = uniformflow(TR = k * (top - base), gradient = 0.002, angle = -135) # SW direction
rf = constant(x = 1000, y = 1000, h = 25)

# create and solve model
m = aem(k, top, base, n, w1, w2, as, uf, rf)

# output grid
xg = seq(-1200, 1000, length = 100)
yg = seq(-700, 500, length = 100)

# plot head contours
contours(m, xg, yg, col = 'dodgerblue', nlevels = 20, labcex = 0.8, xlab = 'x (m)', ylab = 'y (m)')

```

# Exporting as spatial vector data

The contours can be exported as a spatial vector format using the [sf](https://r-spatial.github.io/sf/) and [isoband](https://isoband.r-lib.org/) packages. `sf` is widely used in the R community to handle spatial vector data. `isoband` is a lightweight package containing fast contouring algorithms used in i.a. [ggplot2](https://ggplot2.tidyverse.org/). A contouring grid of the variable of interest is required. Optionally, the origin of the `raem` model grid at $(x, y) = (0,0)$ can be specified in a projected coordinate system so that the exported contours are georeferenced. The set of contour lines is generated with `isoband::isolines()` and converted to a `sf` simple feature geometry collection, which can then be used as a simple feature geometry column in a `sf` simple feature object.

```{r export-vector, message = FALSE}
library(sf)
library(isoband)

# create a 10 by 10 m contouring grid and get the heads as a grid
xg = seq(-1200, 1000, by = 10)
yg = seq(-700, 500, by = 10)

h = heads(m, xg, yg, as.grid = TRUE)

# optionally, set the x and y origin corresponding to (0, 0) 
# in the requested coordinate system
xorigin = 195600
yorigin = 203500
epsg = 31370

# create the isolines with the specified levels
# the y-coordinates need to be reversed here for isolines()
lvls = seq(13.5, 24.5, by = 0.5)
isolines = isolines(xg + xorigin, 
                    rev(yg) + yorigin,
                    h,
                    levels = lvls)

# convert to sfg object, create sfc column for sf object
isolines_sf = st_sf(level = as.numeric(names(isolines)),
                    geometry = st_sfc(iso_to_sfg(isolines)),
                    crs = epsg)

plot(isolines_sf)

```

To export the `sf` object to disk, the `sf` package provides the `write_sf()` function:

```{r export-shapefile, eval = FALSE}
# export as shapefile
write_sf(isolines_sf, 'isolines.shp')
```

# Exporting as spatial raster

To create and export output as a spatial raster object such as a GeoTIFF, the [terra](https://rspatial.github.io/terra/) package can be used. This package is the replacement of the older, widely popular [raster](https://github.com/rspatial/raster) package. Similar to the creation of a spatial vector object, a grid of values for the variable of interest (here hydraulic head) needs to be specified. Optionally, the origin coordinates can be given to georeference the output. The extent of the raster needs to be specified explicitly from the range of the marginal grid vectors.

```{r terra, message = FALSE}
library(terra)

# set extent and create raster
extent = c(range(xg) + xorigin, range(yg) + yorigin)
r = rast(h, crs = paste('epsg', epsg, sep = ':'), extent = ext(extent))

# plot
plot(r)
```

To export the raster to disk, use the `writeRaster()` function from the `terra` package:

```{r write-raster, eval = FALSE}
writeRaster(r, 'heads.tiff', overwrite = TRUE)
```
